---
title: "Check MTA predictions"
output: html_notebook
---

```{r}
library(my.utils)
library(gembox)
library(ComplexHeatmap)
library(venn)
library(rio)
```

```{r}
mta.res <- readRDS("../GEM/mta.res.RDS")
load("check.mta.res.RData")
```


### Comparison of top MTA predicted target reactions between pairs of datasets

```{r}
tmpf <- function(a,b) mapply(function(i,j) enrich.test(top10.mta.hits[[i]]$reaction, top10.mta.hits[[j]]$reaction, recon1$rxns, alternative="greater", confus.mat=NULL)$estimate, a, b)
tmpf1 <- function(a,b) mapply(function(i,j) enrich.test(top10.mta.hits[[i]]$reaction, top10.mta.hits[[j]]$reaction, recon1$rxns, alternative="greater", confus.mat=NULL)$p.value, a, b)
ds <- names(mta.res)
names(ds) <- ds
or.mat <- outer(ds, ds, tmpf)
p.mat <- outer(ds, ds, tmpf1)
p.mat[upper.tri(p.mat)] <- p.adjust(p.mat[upper.tri(p.mat)], "BH")
p.mat[lower.tri(p.mat)] <- p.adjust(p.mat[lower.tri(p.mat)], "BH")
```

```{r}
plot.mat <- function(mat, pmat) {
  tmp <- mat
  tmp[tmp==0] <- min(tmp[tmp!=0])
  cl <- hclust(as.dist(1/tmp))
  mat <- mat[cl$order, ][, cl$order]
  pmat <- pmat[cl$order, ][, cl$order]
  tmp <- mat
  tmp[tmp==0] <- min(tmp[tmp!=0])
  cl <- hclust(as.dist(1/tmp))
  diag(mat) <- NA
  diag(pmat) <- 0
  pmat <- -log10(pmat)
  colf <- circlize::colorRamp2(c(0, -log10(0.05), median(pmat[is.finite(pmat)]), max(pmat[is.finite(pmat)])), c("grey70", "blue", "purple", "red"))
  rf <- function(i=NULL,j=NULL,x=NULL) {
    rmax <- 0.5
    rmin <- 0.2
    xmax <- max(mat[is.finite(mat)])
    if (is.null(x)) x <- mat[i,j]
    if (is.na(x)) res <- NA else if (x>=1) res <- rmin + (rmax-rmin)*(x-1)/(xmax-1) else res <- 0.05 + (0.2-0.05)*x
    if (is.na(res)) NA else res*unit(3,"mm")
  }
  cellf <- function(j, i, x, y, width, height, fill) {
    grid.rect(x=x, y=y, width=width, height=height, gp=gpar(col=NA, fill=NA))
    #grid.circle(x=x, y=y, r=mat[i,j]/max(mat,na.rm=TRUE)/2*min(unit.c(width,height)), gp=gpar(fill=colf(pmat[i,j]), col=NA))
    if (is.infinite(mat[i,j])) grid.circle(x=x, y=y, r=rf(x=max(mat[is.finite(mat)])), gp=gpar(fill=colf(pmat[i,j]), col="black"))
    else grid.circle(x=x, y=y, r=rf(i,j), gp=gpar(fill=colf(pmat[i,j]), col=NA))
  }
  hm <- Heatmap(pmat,
    #name="-log10(padj)",
    width=ncol(pmat)*unit(4,"mm"),
    height=nrow(pmat)*unit(4,"mm"),
    row_title=NULL,
    #row_title_gp=gpar(fontsize=9),
    row_names_gp=gpar(fontsize=8),
    row_names_side="left",
    cluster_rows=cl,
    row_dend_side="right",
    row_dend_width=unit(6,"mm"),
    column_title=NULL,
    #column_title_gp=gpar(fontsize=9),
    column_title_side="bottom",
    column_names_gp=gpar(fontsize=8),
    column_names_rot=40,
    cluster_columns=cl,
    column_dend_height=unit(6,"mm"),
    col=colf,
    rect_gp=gpar(type="none"),
    cell_fun=cellf,
    show_heatmap_legend=FALSE
    #heatmap_legend_param=list(title_gp=gpar(fontsize=6),
    #                          labels_gp=gpar(fontsize=6),
    #                          grid_width=unit(3,"mm"))
  )
  lgd1 <- Legend(col_fun=colf, title=expression(-log[10]*P[adj]), title_gp=gpar(fontsize=8, fontface="plain"), labels_gp=gpar(fontsize=7, fontface="plain"))
  lgd2 <- Legend(title="Odds Ratio", labels=c(15,8,1),
    title_gp=gpar(fontsize=8, fontface="plain"), labels_gp=gpar(fontsize=7, fontface="plain"),
    graphics=list(function(x, y, w, h) grid.circle(x, y, r=rf(x=15), gp=gpar(fill="grey", col=NA)), # requires the dev version of ComplexHeatmap
                  function(x, y, w, h) grid.circle(x, y, r=rf(x=8), gp=gpar(fill="grey", col=NA)),
                  function(x, y, w, h) grid.circle(x, y, r=rf(x=1), gp=gpar(fill="grey", col=NA)))
  )
  lgd <- packLegend(lgd1, lgd2)
  draw(hm, annotation_legend_list=lgd)
}
```

```{r, fig.width=2.35, fig.height=1.5, warning=FALSE}
plot.mat(or.mat, p.mat)
```


### Several validations

With CRISPR-Cas9 screen hits:

```{r}
rbindlist(lapply(enr.res10.rxn, function(x) x$screen[reaction.set=="pos.crispr.union"]), idcol="dataset")[, padj:=p.adjust(pval,"BH")][order(padj,pval), -"reaction.set"]
rbindlist(lapply(enr.res20.rxn, function(x) x$screen[reaction.set=="pos.crispr.union"]), idcol="dataset")[, padj:=p.adjust(pval,"BH")][order(padj,pval), -"reaction.set"]
```

Visualization:

```{r, fig.width=2.3, fig.height=1.4}
tmp <- rbindlist(lapply(enr.res10.rxn, function(x) x$screen[reaction.set=="pos.crispr.union"]), idcol="dataset")[, padj:=p.adjust(pval,"BH")][order(padj,pval), -"reaction.set"]
tmp[, dataset:=factor(dataset, levels=dataset)]
tmp[, nlp:=-log10(padj)]

p <- ggplot(tmp) +
  ylab("Odds Ratio") +
  geom_point(aes(x=dataset, y=odds.ratio, color=nlp, size=overlap.size)) + 
  scale_color_continuous(low="grey", high="red3", name=expression(-log[10]*P[adj])) + 
  scale_size_continuous(name="overlap size", range=c(1,4)) +
  geom_hline(yintercept=1, linetype="dashed", color="grey") +
  annotate("text", x=1.7, y=1.4, label="odds ratio>1", size=2.8, color="grey30") +
  theme_classic() +
  thm(y.tit=10, y.txt=8, x.tit=NULL, x.txt=list(size=8,angle=30,hjust=1), lgd.tit=8, lgd.txt=6, lgd.key=10) + theme(legend.spacing=unit(0,"pt"))
p
```

AUROC based on CRISPR-Cas9 screens:

```{r}
rbindlist(lapply(auc.res, function(x) x$crispr.union), idcol="dataset")[order(-auroc)]
```

Visualization:

```{r, message=FALSE, fig.width=1.6, fig.height=1}
# recompute AUROC data with 95% CI for plotting
tmp <- rbindlist(lapply(mta.res, function(x) {
  tmp <- get.roc1(x[!is.na(score),score], pos=rxns.screen.for.roc$pos.crispr.union, neg=rxns.screen.for.roc$neg.crispr.union, x.names=x[!is.na(score),reaction], msg=FALSE)
  data.table(auc=tmp$auc, lb=tmp$auc.ci[1], ub=tmp$auc.ci[2])
}), idcol="dataset")[order(-auc)]
tmp[, dataset:=factor(dataset, levels=dataset)]

p <- ggplot(tmp) + scale_y_continuous(name="AUROC", breaks=c(0.3,0.5,0.7,0.9)) +
  geom_pointrange(aes(x=dataset, y=auc, ymin=lb, ymax=ub, color=auc)) +
  scale_color_gradient2(low="grey", mid="grey", high="red3", midpoint=0.5) +
  geom_hline(yintercept=0.5, linetype="dashed", color="grey40") +
  theme_classic() +
  thm(x.tit=NULL, x.txt=list(size=8, angle=35, hjust=1), y.tit=10, y.txt=8, lgd="none")
p
```

Examine some ROC curves:

```{r}
# plot ROC curves
plot.roc1 <- function(ds, ci=TRUE) {
  dat <- get.roc1(mta.res[[ds]][!is.na(top.percent),100-top.percent], pos=rxns.screen.for.roc$pos.crispr.union, neg=rxns.screen.for.roc$neg.crispr.union, x.names=mta.res[[ds]][!is.na(top.percent),reaction], ci=ci)
  plot.roc(dat, col=function(x) 100-x, rev.lgd=TRUE, lgd.tit="MTA\nTop%", lab.size=2.8) + ggtitle(ds) + thm(x.tit=10, y.tit=10, x.txt=8, y.txt=8, tit=10, lgd.tit=8, lgd.txt=8, lgd.key=10)
}
```

```{r, fig.width=1.35, fig.height=1.1}
plot.roc1("Calu-3")
plot.roc1("SC.Liao")
```

With experimentally validated drugs (union of those compiled by Kuleshov et al.):

```{r}
rbindlist(lapply(enr.res10.rxn, function(x) x$drugs.exp[reaction.set=="union"]), idcol="dataset")[, padj:=p.adjust(pval,"BH")][order(padj,pval), -"reaction.set"]
rbindlist(lapply(enr.res20.rxn, function(x) x$drugs.exp[reaction.set=="union"]), idcol="dataset")[, padj:=p.adjust(pval,"BH")][order(padj,pval), -"reaction.set"]
rbindlist(lapply(gsea.res, function(x) x$drugs.exp[pathway=="union"]), idcol="dataset")[, padj:=p.adjust(pval,"BH")][order(padj,pval), -"pathway"]
```

Any significantly enriched experimentally validated drug set:

```{r}
rbindlist(lapply(enr.res10.rxn, function(x) x$drugs.exp[padj<0.2]), idcol="dataset")
rbindlist(lapply(enr.res20.rxn, function(x) x$drugs.exp[padj<0.2]), idcol="dataset")
```

Visualization:

```{r, fig.width=4.4, fig.height=1.1}
tmp <- rbindlist(lapply(enr.res10.rxn, function(x) x$drugs.exp[padj<0.1]), idcol="dataset")[order(padj,pval)]
tmp[, ds:=paste0(dataset, ": ", reaction.set)]
tmp[, ds:=factor(ds, levels=rev(ds))]
p <- ggplot(tmp, aes(x=odds.ratio, y=ds)) +
  xlab("Odds Ratio") +
  geom_point(aes(color=padj, size=overlap.size)) + 
  scale_color_continuous(low="red3", high="grey", name=expression(P[adj]), guide=guide_colorbar(reverse=TRUE)) + 
  scale_size_continuous(name="overlap size", range=c(1,4)) +
  geom_vline(xintercept=1, linetype="dashed", color="grey") +
  annotate("text", x=2, y=7, label="odds ratio>1", size=2.8, color="grey30", angle=270) +
  theme_classic() +
  thm(x.tit=9, x.txt=8, y.tit=NULL, y.txt=8, lgd.tit=8, lgd.txt=6, lgd.key=10, plt.margin=c(1,0,0,0)) + theme(legend.spacing=unit(0,"pt"))
p
```

Viral PPI genes:

```{r, fig.width=2.6, fig.height=1}
tmp <- rbindlist(lapply(enr.res10.rxn, function(x) x$ppi[reaction.set %in% c("ppi.gordon","ppi.stukalov")]), idcol="dataset")[, padj:=p.adjust(pval,"BH")][padj<0.1][order(padj,pval)]
tmp[, ds:=paste0(dataset, ": host-viral PPI from Stukalov et al.")]
tmp[, ds:=factor(ds, levels=rev(ds))]
p <- ggplot(tmp, aes(x=odds.ratio, y=ds)) +
  xlab("Odds Ratio") + xlim(c(1.45,1.55)) +
  geom_point(aes(color=pval, size=overlap.size)) + 
  scale_size_continuous(name="overlap size", breaks=c(44,46), range=c(2,4)) +
  scale_color_continuous(low="red3", high="grey", name="P value", guide=guide_colorbar(reverse=TRUE)) + 
  theme_classic() +
  thm(x.tit=9, x.txt=8, y.tit=NULL, y.txt=8, lgd.tit=8, lgd.txt=6, lgd.key=10, plt.margin=c(1,0,0,0)) + theme(legend.spacing=unit(0,"pt"))
p
```

### Get final list of candidate targets

Check some Venn diagrams:

```{r, fig.width=2, fig.height=2}
tmp <- lapply(top10.mta.hits[c("Swab.Butler","Swab.Lieberman","BALF","SC.Liao","SC.Chua.Basal","SC.Chua.Ciliated")], function(x) x$reaction)
venn(tmp, zcolor="style", ilcs=1, sncs=1)
tmp <- lapply(top10.mta.hits[c("NHBE","Vero","Calu-3","A549","293T","Caco-2")], function(x) x$reaction)
venn(tmp, zcolor="style", ilcs=1, sncs=1)
```

Final list: the intersection of top 10% of at least 3 from {in vivo} and at least 3 from {in vitro}

```{r}
hits <- intersect(
  rbindlist(top10.mta.hits[c("Swab.Butler","Swab.Lieberman","BALF","SC.Liao","SC.Chua.Basal","SC.Chua.Ciliated")])[, .(n=.N), by=reaction][n>=3, reaction],
  rbindlist(top10.mta.hits[c("NHBE","Vero","Calu-3","A549","293T","Caco-2")])[, .(n=.N), by=reaction][n>=3, reaction]
)

length(hits) #hit rxns
bg.rxns <- recon1$rxns[recon1$rules!=""] 
sum(hits %in% bg.rxns) #hit rxns with mapped genes
uniqueN(unlist(rxns2genes(recon1, hits))) #mapped genes
sum(sapply(rxns2genes(recon1, hits), function(x) any(d2t$target %in% x))) #rxns with drugs
d2t[target %in% unlist(rxns2genes(recon1, hits))][, uniqueN(target)] #genes with drugs
d2t[target %in% unlist(rxns2genes(recon1, hits))][, uniqueN(drug)] #drugs
```

Balidation of final hits:

```{r}
enrich.gsets(hits, vrxns$screen, recon1$rxns)
enrich.gsets(hits, vrxns$ppi, recon1$rxns)
enrich.gsets(hits, vrxns$drugs.exp, recon1$rxns)
lapply(screen.full, function(x) gsea(x, list(hits=unique(unlist(rxns2genes(recon1,hits))))))
```

Pathway enrichments of final hits:

```{r}
rset <- subsystems2gsets(recon1)
enrich.gsets(hits, rset, recon1$rxns)[padj<0.1]
tx.ce <- get.transport.info(recon1)
enrich.gsets(hits, lapply(tx.ce, function(x) x$rxn), recon1$rxns)[padj<0.1]
```

```{r, fig.width=2.1, fig.height=1.2}
tmp <- rbind(enrich.gsets(hits, rset, recon1$rxns)[padj<0.1], NA, enrich.gsets(hits, lapply(tx.ce, function(x) x$rxn), recon1$rxns)[padj<0.1 & !gene.set %in% c("na1","cl","nh4","hco3","k","h2o2","ca2")][, gene.set:=recon1$metNames[all2idx(recon1, paste0(gene.set,"[c]"))]], fill=TRUE)
tmp[is.na(gene.set), gene.set:=""]
tmp[, gene.set:=factor(gene.set, levels=gene.set)]
tmp[, nlp:=-log10(padj)]

p <- ggplot(tmp, aes(x=gene.set, y=odds.ratio)) +
  ylab("Odds Ratio") + ylim(c(0,200)) +
  geom_point(aes(color=nlp, size=overlap.size)) +
  scale_size_continuous(name="overlap size", breaks=c(10,20,40), range=c(1,4)) +
  scale_color_continuous(low="grey", high="red3", name=expression(-log[10]*P[adj])) + 
  geom_hline(yintercept=1, linetype="dashed", color="grey") +
  annotate("text", x=5, y=15, label="odds ratio>1", size=2.6, color="grey30") +
  theme_classic() +
  thm(x.tit=NULL, x.txt=list(size=9,angle=30,hjust=1), y.tit=10, y.txt=9, lgd="bottom", lgd.tit=8, lgd.txt=6, lgd.key=10, plt.margin=c(0,0,0,50), lgd.margin=c(-30,10,0,0)) + theme(legend.spacing=unit(0,"pt"))
p
```

Export tables:

```{r}
hits <- intersect(
  rbindlist(top10.mta.hits[c("Swab.Butler","Swab.Lieberman","BALF","SC.Liao","SC.Chua.Basal","SC.Chua.Ciliated")])[, .(n=.N), by=reaction][n>=3, reaction],
  rbindlist(top10.mta.hits[c("NHBE","Vero","Calu-3","A549","293T","Caco-2")])[, .(n=.N), by=reaction][n>=3, reaction]
)
tps <- sapply(mta.res, function(x) x[match(hits,reaction), top.percent])
use <- tps<20
tps[!use] <- 20
res <- data.table(reaction=hits, average.top.percent=rowMeans(tps), datasets=apply(use, 1, function(x) cc(names(x)[x])))
res <- res[order(average.top.percent)]
res[, pathway:=recon1$subSystems[all2idx(recon1, reaction)]]
res[, genes:=rxns2genes(recon1, reaction)]
res <- res[sapply(genes, length)!=0]
res[, crispr.hits:=lapply(genes, function(x) x[x %in% vgenes$screen$pos.crispr.union])]
res[, ppi.genes:=lapply(genes, function(x) x[x %in% vgenes$ppi$ppi.union])]
res[, drugs:=lapply(genes, function(x) d2t[target %in% x, unique(drug)])]
res[, drug.hits:=lapply(drugs, function(x) x[x %in% drugs.exp$union$drug])]
res[, drugs:=lapply(genes, function(x) d2t[target %in% x, .(targets=list(target)), by=drug][, co(paste0(drug,":{",sapply(targets, co, c=","),"}"), c="|")])]
res[drugs==":{}", drugs:=""]
saveRDS(res, file="final.mta.hits.RDS")
write.tab(res, "final.mta.hits.tsv", quote=FALSE)
```

```{r}
tmp1 <- lapply(mta.res, function(x) x[, .(reaction, score=norm.score, top.percent, genes, pathway)])
tmp2 <- rbindlist(lapply(enr.res10.rxn, function(x) x$screen[reaction.set=="pos.crispr.union"]), idcol="dataset")[, padj:=p.adjust(pval,"BH")][order(padj,pval), -"reaction.set"]
tmp3 <- rbindlist(lapply(enr.res10.rxn, function(x) x$drugs.exp[padj<0.1]), idcol="dataset")[order(padj,pval)]
tmp4 <- rbindlist(lapply(enr.res10.rxn, function(x) x$ppi[reaction.set %in% c("ppi.gordon","ppi.stukalov")]), idcol="dataset")[, padj:=p.adjust(pval,"BH")][order(padj,pval)]
tmp <- c(tmp1, list(S6B=tmp2, S6C=tmp3, S6D=tmp4))
rio::export(tmp, "Table S6_MTA-predicted anti-SARS-CoV-2 metabolic targets based on each dataset and their validations.xlsx")


tmp <- rbind(enrich.gsets(hits, rset, recon1$rxns, name="reaction"), NA, enrich.gsets(hits, lapply(tx.ce, function(x) x$rxn), recon1$rxns, name="reaction")[, reaction.set:=recon1$metNames[all2idx(recon1, paste0(reaction.set,"[c]"))]], fill=TRUE)
tmp[, x:=NULL]
tmp <- list(S7A=res, S7B=tmp)
rio::export(tmp, "Table S7_Consensus MTA-predicted anti-SARS-CoV-2 metabolic targets across datasets and their pathway enrichment.xlsx")
```




